\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-26 paper.tex
delete dead stuff
Add data availability section

The graph visualisations in Fig.~\ref{fig-simplification} have
three novel features which require some explanation.
Firstly, edge weights (the thickness of the lines joining
nodes) correspond to the length of the inheritance intervals
they are annotated with. This allows us to distinguish
edges that persist across many local trees from those that are
less influential (contrast the edge
$(\noderef{g}, \noderef{h})$
with $(\noderef{g}, \noderef{i})$
in Fig.~\ref{fig-simplification}A).
% FIXME this needs some work
Secondly, node colours denote the number of parents that they
have in the graph, allowing us to easily see roots (those
with zero parents), recombinants (those with two parents)
and more complex situations arising from simplification (see below).
Thirdly, the shading intensity of a node denotes the ``coalescent span'', the
fraction of the node's span (the length of genome in which it
is reachable from the samples in the local trees)
over which it is coalescent (has more than one child). Nodes which
are never locally unary therefore have a coalescent span of 100\%,
whereas nodes in which ancestral material never coalesces have a
coalescent span of 0\%.
% JK: Let's not get into discussions of identifiability or inferability
% here, we're just trying to show the things that exist in the true history.
% (such nodes will be extremely difficult to
% identify from the distribution of shared mutations in the samples).
% This is useful why? I can't come up with a quick way of saying
% this. In some senses it's a measure of identifiability?
% JK: Hmm, not sure we want to jump into inferability here?
% All these features will impact what kind of an ARG we can
% hope to infer from sampled genomes.

@@

To illustrate the varying levels of information that can be captured in a gARG,
Fig.~\ref{fig-inferred-args} presents graphical representations of genealogies output by two tools that
\emph{do} explicitly infer recombination events, namely \kwarg~\citep{ignatieva2021kwarg} and
\argweaver~\citep{rasmussen2014genome},
and two which do not (\tsinfer\ and \relate).
All inferences are based on the
same underlying dataset from \citet{kreitman1983nucleotide}, a standard benchmark in the classical
ARG literature, for which the minimum number of recombination breakpoints in the absence of recurrent
mutation is 7~\citep{song2003parsimonious}. This dataset comprises 43 biallelic SNPs spanning 2.4Kb of the
\emph{D.~melanogaster} ADH locus on chromosome 2L. Where required for inference purposes we
assume a mutation rate of
$5.49\times10^{-9}$ per site per generation
and a recombination rate of
$8.4\times10^{-9}$ per site per generation
\citep{adrion2020community,lauterbur2022expanding},
together with
a constant effective population size of 1,000,000.
To focus on the ARG topology rather than estimates of node times, the Y position of each node
is arbitrary, taken from the ranked order of nodes output by each software method.
The version numbers of inference software, the specific inference parameters
needed to reproduce the ARGs, and (where necessary) functions to convert outputs to the
\texttt{tskit} gARG format, are available from our GitHub repository.

@@

>>>>

The graph visualisations in Fig.~\ref{fig-simplification} have
three novel features which require some explanation.
Firstly, edge weights (the thickness of the lines joining
nodes) correspond to the length of the inheritance intervals
they are annotated with. This allows us to distinguish
edges that persist across many local trees from those that are
less influential (contrast the edge
$(\noderef{g}, \noderef{h})$
with $(\noderef{g}, \noderef{i})$
in Fig.~\ref{fig-simplification}A).
% FIXME this needs some work
Secondly, node colours denote the number of parents that they
have in the graph, allowing us to easily see roots (those
with zero parents), recombinants (those with two parents)
and more complex situations arising from simplification (see below).
Thirdly, the shading intensity of a node denotes the ``coalescent span'', the
fraction of the node's span (the length of genome in which it
is reachable from the samples in the local trees)
over which it is coalescent (has more than one child). Nodes which
are never locally unary therefore have a coalescent span of 100\%,
whereas nodes in which ancestral material never coalesces have a
coalescent span of 0\%.
% JK: Let's not get into discussions of identifiability or inferability
% here, we're just trying to show the things that exist in the true history.
% (such nodes will be extremely difficult to

@@

To illustrate the varying levels of information that can be captured in a gARG,
Fig.~\ref{fig-inferred-args} presents graphical representations of genealogies output by two tools that
\emph{do} explicitly infer recombination events, namely \kwarg~\citep{ignatieva2021kwarg} and
\argweaver~\citep{rasmussen2014genome},
and two which do not (\tsinfer\ and \relate).
All inferences are based on the
same underlying dataset from \citet{kreitman1983nucleotide}, a standard benchmark in the classical
ARG literature, for which the minimum number of recombination breakpoints in the absence of recurrent
mutation is 7~\citep{song2003parsimonious}. This dataset comprises 43 biallelic SNPs spanning 2.4Kb of the
\emph{D.~melanogaster} ADH locus on chromosome 2L. Where required for inference purposes we
assume a mutation rate of
$5.49\times10^{-9}$ per site per generation
and a recombination rate of
$8.4\times10^{-9}$ per site per generation
\citep{adrion2020community,lauterbur2022expanding},
together with
a constant effective population size of 1,000,000.
To focus on the ARG topology rather than estimates of node times, the Y position of each node
is arbitrary, taken from the ranked order of nodes output by each software method (see the ``Data Availability''
section for details of inference software and conversion methods).

@@

\section*{Data Availability}

The public GitHub repository at \url{https://github.com/tskit-dev/what-is-an-arg-paper} can be
used to reproduce all figures and tables in this paper. In particular this includes
the ARG used in Fig.~\ref{fig-ancestry-resolution}; the simulation code and functions
used to generate Fig.~\ref{fig-simplification}; and for Fig.~\ref{fig-inferred-args},
the software versions, parameter settings, and (where necessary) functions to convert
software outputs to the \texttt{tskit} gARG format.
